---
title: "Measures of Dispersion"
author: "Hafeez Nathoo"
date: "2024-07-23"
categories: [ statistics, analysis, data]
---

The goal of this post is to give a quick introduction to measures of dispersion - which is a fancy way of saying, how can we use a single number to summarize the contents of a data set. The four quantities we are concerned with will be the **standard deviation**, **variance**, **inter quartile range (IQR)**, and **range**.

The big question to answer is why do we care? 
Generally, data can be described with two measures - one of [central tendency]<https://mathformedstudents.netlify.app/topics/2_central_tendency/> and one of 

Lets start with a simple data set for illustrative purposes with only five numbers: $(1, 4, 5, 10, 10)$

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
sample <- c(1, 4, 5, 10, 10)

sd(sample)
var(sample)
IQR(sample)
range <- range(sample)

data_range <- range[2] - range[1]
```


```{r echo=FALSE, messages=FALSE, warning=FALSE, error=FALSE}

normalvec <- rnorm(1000)

ggplot(mapping = aes(normalvec))+
  geom_histogram(color = "black", fill = "steelblue4", bins = 40)+
  # geom_vline(xintercept = mean(normalvec), color = 'black')+
  # geom_vline(xintercept = median(normalvec), color = 'red')+
  xlab("Values")+
  ylab("count")+
  theme_bw()

```

Ultimately the best measure to report will depend on you application - these measures are the first step to deeper statistical analysis.


### Thank you!

Hope the above was helpful, a few concepts were touched on in this article - skew, data visualization, and so on - stay tuned for more posts diving deeper into those topics!


## 

**References**

1.  *El Omda S, Sergent SR. Standard Deviation. [Updated 2023 Aug 14]. In: StatPearls [Internet]. Treasure Island (FL): StatPearls Publishing; 2024 Jan-. Available from: https://www.ncbi.nlm.nih.gov/books/NBK574574/*
2.  *Wadhwa RR, Azzam D. Variance. [Updated 2023 Jul 17]. In: StatPearls [Internet]. Treasure Island (FL): StatPearls Publishing; 2024 Jan-. Available from: https://www.ncbi.nlm.nih.gov/books/NBK551689/*

