---
title: "Measures of Central Tendency"
author: "Hafeez Nathoo"
date: "2024-06-24"
categories: [StatisticsForMedStudents]
---

The goal of this post is to give a quick introduction to measures of central tendency - which is a fancy way of saying, how can we use a single number to summarize the contents of a data set. The three quantities we are concerned with will be the **mean (average)**, **median**, and **mode**.

These are widely taught in pretty much every stats class, so here we will focus on conceptual meaning and interpretation.

Lets start with a simple data set for illustrative purposes with only five numbers: $(1, 4, 5, 10, 10)$

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
sample <- c(1, 4, 5, 10, 10)
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
# 
# mean(sample)
# median(sample)
# getmode(sample)
```

### Mean (aka the Average)

The mean ($\mu$) is defined as the sum of all members of the set ($x_i$'s) divided by the number of members in that set ($n$) - this works out to $30/5$ for our example giving a mean of $6$.

$$
\mu = \frac{\sum_{i}^{n} x_i}{n}
$$
    
### Median

The median is defined as the middle value in the set, also known as the 50th percentile. This value will represent the halfway point - half the numbers are less, and half more.

In our data set $(1, 4, 5, 10, 10)$, the median is $5$.

If there were two middle numbers - for example in $(2, 3, 4, 5, 6, 8)$ - then average the two middle numbers. For this example, the median is $4.5$.


### Mode

The mode is defined as the value which appears most frequently in the data set $(1, 4, 5, 10, 10)$, this is straightforward and is clearly the value $10$ in our sample data.

### The Importance of Mean vs Median

These definitions are straightforward enough, and are often calculated with software (R, Excel, python, etc.) and not by hand. The importance is in understanding their utility - in particular Mean vs Median.

The mean can be affected by outliers - e.g. the mean of $(1,2,3,4,100)$ is $22$ - a high value considering the bulk of the data is below 5. The median is $3$ - much more representative of the bulk of the data. The median is commonly reported in clinical applications owing to its resistance to the effects of outliers.

In a normal (bell) curve, the mean and median are fairly similar - as expected when looking at the histogram below. 
<!-- The mean is the black line and median is in red. -->

```{r echo=FALSE, messages=FALSE, warning=FALSE, error=FALSE}

normalvec <- rnorm(1000)

ggplot(mapping = aes(normalvec))+
  geom_histogram(color = "black", fill = "steelblue4", bins = 40)+
  # geom_vline(xintercept = mean(normalvec), color = 'black')+
  # geom_vline(xintercept = median(normalvec), color = 'red')+
  xlab("Values")+
  ylab("count")+
  theme_bw()

```

When you have data which is not as close to a bell curve (demonstrated in the below histogram) - median can be more useful since it is not as susceptible to the effects of a long tail (also known as **skew**). 
<!-- The mean is the black line and median is in red. -->
In the below data, the extreme values to the right would inflate the value of the mean.

```{r echo=FALSE, messages=FALSE, warning=FALSE}

lognormalvec <- rlnorm(5000)

ggplot(mapping = aes(lognormalvec))+
  geom_histogram(color = "black", fill = "steelblue4", bins = 60)+
  # geom_vline(xintercept = mean(lognormalvec), color = 'black')+
  # geom_vline(xintercept = median(lognormalvec), color = 'red')+
  xlab("Values")+
  ylab("count")+
  theme_bw()
```

Ultimately the best measure to report will depend on your application - these measures are the first step to deeper statistical analysis.


### Thank you!

Hope the above was helpful, a few concepts were touched on in this article - skew, data visualization, and so on - stay tuned for more posts diving deeper into those topics!





## 

**References**

1.  *Hurley M, Tenny S. Mean. \[Updated 2023 Jul 17\]. In: StatPearls \[Internet\]. Treasure Island (FL): StatPearls Publishing; 2024 Jan-.Available from: https://www.ncbi.nlm.nih.gov/books/NBK546702*
2.  *Tenny S, Hoffman MR. Median. \[Updated 2022 Sep 19\]. In: StatPearls \[Internet\]. Treasure Island (FL): StatPearls Publishing; 2024 Jan-.Available from: https://www.ncbi.nlm.nih.gov/books/NBK470533*
3.  *Mercadante AA, Tenny S. Mode. \[Updated 2023 Jul 17\]. In: StatPearls \[Internet\]. Treasure Island (FL): StatPearls Publishing; 2024 Jan-.Available from: https://www.ncbi.nlm.nih.gov/books/NBK540990*
